---
title: "chapter6"
author: "Feroze"
date: "8 12 2019"
output: html_document
---

# 6 Longtitudinal data {.tabset}

this week exercises is based on the Chapters 8 and 9 of Vehkalahti and Everitt (2019), included in the special edition MABS4IODS (Part VI) we work with two dataset, BPRS and RATS, and perform the analyses based on swapping method.

## first dataset RATS

This dataset RATS has 16 rats for study, and they are divided into 3 groups based on different diets.

```{r}
library(ggplot2)
library(tidyr)
library(dplyr)
RATSL <- read.csv(file = "C:/Users/ferozef/Documents/R/IODS-project/data/RATSL.csv", row.names = 1)

# categorials into factors
RATSL$Group <- factor(RATSL$Group)
RATSL$ID <- factor(RATSL$ID)
# stardardize bprsl
RATSL <- RATSL %>% group_by(Time) %>% mutate(stweight = (Weight - mean(Weight)) / sd(Weight)) %>% ungroup()
str(RATSL)
```

## Graphical exploration of longtitudinal data

```{r}
# plot the individuals in time with standardized bprs value 
ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID)) + geom_line() + facet_grid(. ~ Group, labeller = label_both) + scale_linetype_manual(values = rep(1:6, times=4)) + theme(legend.position = "none") +   scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))
```

The diet has influenced the rats wieght and all the rats has shown to gain weight. Comparing all the 3 group, Group1 rats are at the baseline whereas, Group 2 and 3 has higher weight and variability compared to Group1

```{r}
# Summary plot
# Number of timepoints, baseline included
n <- RATSL$Time %>% unique() %>% length()
# Summarize mean and std error
rats_sum <- RATSL %>% group_by(Group, Time) %>% summarise(mean = mean(Weight), se = ( sd(Weight) / sqrt(n) )) %>% ungroup()
glimpse(rats_sum)
# plot the plots
par(mfrow = c(1,2))
# with means only
ggplot(rats_sum, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  theme(legend.position = c(0.9,0.5)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
# added with the error bar
ggplot(rats_sum, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.9,0.5)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
```
from the above graph, we can conclude that all the 3 groups weight increase by time. 


```{r}
# boxplot
ggplot(RATSL, aes(x = factor(Time), y = Weight, fill = Group)) + geom_boxplot()
 ```

Boxplots with non-summarized data.

## Preparinf the summary measure approach

```{r}
# Summary measure approach, filter out outliers
# create the summary dataset with means from the weeks 1-8
rat8s <- RATSL %>%
  filter(Time > 0) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()
glimpse(rat8s)
ggplot(rat8s, aes(x=Group, y = mean)) + geom_boxplot() + stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white")
```
From the above boxplot, we can clearly see the outliers present in the group, let's filter out the outliers from the data.

```{r}
# filtering out the outlier's and re-plotting the box plot again

rat8s1 <- filter(rat8s, Group == 2 & mean < 550 | Group == 1 & mean > 260 | Group == 3 & mean > 500)
glimpse(rat8s1)
ggplot(rat8s1, aes(x=Group, y = mean)) + geom_boxplot() + stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white")
```
From the above box plot 2, we can see that the outlier's has been filtered out and the data now seems to be good. 



## Summary approach, comparing means

```{r}
summary(aov(mean ~ Group, data = rat8s1))
```

T-test to the summary data without the outliers, shows no difference between the groups (p=.60, conf.interval includes zero).


```{r}
kruskal.test(mean ~ Group, data = rat8s1)
```

The differences between the group are statistically significant with p-value = 0.007119 according to both tests.

```{r}
# covariance analysis with the baseline
# add the baseline from the original wide dataset to the summary data
rats <- read.csv(file = "C:/Users/ferozef/Documents/R/IODS-project/data/RATS.csv", row.names = 1)
rat8s2 <- rat8s %>% mutate(baseline = rats$WD1)
glimpse(rat8s2)
# Fittinf linear regression model
fit <- lm(mean ~ baseline + Group, data = rat8s2)
summary(fit)
```

The covariance analysis reveals, that the baseline is highly related to the means with 5.02e-08 ***.
And, after controlling the baseline, there is no significant diffence left between the group means.

## BPRS

Do the growth profiles of the treatment groups differ?

```{r}
# read the data, in long form 
bprsl <- read.csv(file = "C:/Users/ferozef/Documents/R/IODS-project/data/BPRSL.csv", row.names = 1)
bprsl$treatment <- factor(bprsl$treatment)
bprsl$subject <- factor(bprsl$subject)
glimpse(bprsl)
```

```{r}
ggplot(bprsl, aes(x = weeks, y = bprs, group = subject)) + geom_line() + facet_grid(. ~ treatment, labeller = label_both) 
```
We have the individuals, all with bprs points decreasing in time, some more than others.


```{r}
bps_reg <- lm(bprs ~ weeks + treatment, data=bprsl)
summary(bps_reg)
```


### Random intercept model
```{r}
library(lme4)
bps_ref <- lmer(bprs ~ weeks + treatment + (1|subject), data = bprsl, REML=FALSE)
summary(bps_ref)
```

Note that in this model the within-subject dependencies are ignored. See std.dev of subject.

### Random intercept and random slope model 

```{r}
bps_ref2 <- lmer(bprs ~ weeks + treatment + (weeks|subject), data = bprsl, REML=FALSE)
summary(bps_ref2)
```

```{r}
anova(bps_ref, bps_ref2)
```

The model fit is better in model 2.

### Random intercept and random slope model with interaction

```{r}
bps_ref3 <- lmer(bprs ~ weeks * treatment + (weeks|subject), data = bprsl, REML=FALSE)
summary(bps_ref3)
```

The estimated regression parameters for the interaction  indicate that the bprs decrease slopes are somewhat higher for individuals in treatment group 2 than in group 1. 


```{r}
anova(bps_ref2, bps_ref3)
```

The fit is not significantly better in model 3.


```{r}
# Create a vector of the fitted values
Fitted <- fitted(bps_ref2)
bprsl <- bprsl %>% mutate(Fitted)
# draw the plot of BPRSL with the observed bprs values
p1 <- ggplot(bprsl, aes(x = weeks, y = bprs, group = subject)) + geom_line() + facet_grid(. ~ treatment, labeller = label_both) + scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 8, 1)) +  ylim(0,100) + theme(legend.position = "top") + ggtitle("Observed")
# draw the plot with fitted
p2 <- ggplot(bprsl, aes(x = weeks, y = Fitted, group = subject)) +
  geom_line() + facet_grid(. ~ treatment, labeller = label_both) +
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 8, 1)) +
  ylim(0, 100) +
  theme(legend.position = "top") + ggtitle("Fitted")
p1; p2
```
