---
title: "chapter4"
author: "Feroze"
date: "11/24/2019"
output: html_document
---

# Chapter 4: Classification and clustering {.tabset}


 I will analyze Boston data set available in MASS package. 


setwd("C:/Users/ferozef/Documents/R/IODS-project")

 setwd("C:/Users/Fariah/Documents/IODS-project")

###More information on Boston data set:  

<https://stat.ethz.ch/R-manual/R-devel/library/MASS/html/Boston.html>

##Load Boston data set

Boston data set contains 506 housing observation (rows) with 14 variables (columns).

```{r, message=FALSE}
library(MASS)
data('Boston')
str(Boston)
summary(Boston)
```

## Visualize and summarize data

Nice corrplot modification examples <https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html>

```{r, message=FALSE}
library(corrplot)

correlation <- cor(Boston)
round(correlation, digits = 2)
corrplot(correlation, method="circle")
corrplot(correlation, type = 'upper')
```

 negative correlation in red and positive correlation in blue. Narrowness of `method = 'circle'` Darkness/weakness of colour indicates how high/low correlation are respectively.  

From the graph it can be observed that most of the variables correlate significantly with other variables. Only few pairs are not significantly correlated.

## Scaling and categorising the variables


To classify our data, variable values need to be standarized/scaled so that all variables have a mean value to zero `0´. It is done as follows (when all variables are numerical, as expected for classification analysis):

```{r}
scaled <- as.data.frame(scale(Boston))
class(scaled)
str(scaled)
summary(scaled)
```
`crim` is categorised according to quantiles and later will be used to train the model to predict the right crime rate class of an observation based on other variables.

```{r}
bins = quantile(scaled$crim)
crime <- cut(scaled$crim, breaks = bins, label = c('low', 'med_low', 'med_high', 'high'), include.lowest = TRUE)
#count table for each category level
table(crime)
#replace original crim with categorical crime variable
scaled <- dplyr::select(scaled, -crim)
scaled <- data.frame(scaled, crime)
head(scaled)
```
## Divide data for training and test sets

To demonstrate, how well my model is predicting crime rate, I will use the cut-off values (20%) proportion of the variable data for testing it, so that 80% of the data belongs to the train set. Observations are selected randomly below for training or test sets.

```{r}
random_test_rows <- sample(nrow(scaled), size = nrow(scaled) * 0.2)
test_set <- scaled[random_test_rows, ]
train_set <- scaled[-random_test_rows, ]
#Check that resulting dfs are as should
dim(test_set)
dim(train_set)
```



## Fitting linear discriminant analysis for crimes

I used lda() function for Fitting classification model using crimes as a categorical variable and all other (continuous) variables as predicting variables.

```{r}
lda_fit <- lda(crime ~ ., data = train_set)
lda_fit
```
LD1 has the highest proportions of variable as its predictors, as evident from the results.

## Visualization of model


```{r, message=FALSE, warning=FALSE}
#Convert crime factor levels to numeric to plot in different colors
crime_levels <- as.numeric(train_set$crime)
#I hate the default colors of the plot, so I'm using ggsci package palettes instead
#Good source for color palettes https://www.datanovia.com/en/blog/top-r-color-palettes-to-know-for-great-data-visualization/
library(ggsci)
library(devtools)
install_github("fawda123/ggord")
library(ggord)
cols <- pal_aaas()(4)
#Plot with nicer colors
ggord(lda_fit, train_set$crime, poly = FALSE, arrow=.3, veclsz = .5, vec_ext = 4, size=1, cols = cols)
```



## Test model

I created corrected crime classes for the test data and removed crime variable from test data that is used to predict the classes with the `lda_fit`.

```{r}
#Save correct classes to variable
correct_classes <- test_set$crime
#Remove classes from test data
test_set <- dplyr::select(test_set, -crime)
#Predict classes with model
lda_predict <- predict(lda_fit, newdata = test_set)
#Make 2X2 table to observe model accuracy
table(correct = correct_classes, predicted = lda_predict$class)
#Calculate percentage of right predictions on test data
percent_correct <- 100 * mean(lda_predict$class==correct_classes)
percent_correct <- round(percent_correct, digits = 0)
percent_correct
```

The table shows the actual and predicted categories. The model predicts best the category of highs  - aligned with the graphical observation above. With other categories there are errors, but majority is correct. 

## K-means and distances 

```{r}
# re-data loading again and standardize it
data(Boston)
boston_sc <- scale(Boston)
# Compute distances (Euclidean is the default)
dist_eu <- dist(boston_sc)
dist_man <- dist(boston_sc, method="manhattan")
summary(dist_eu)
summary(dist_man)
```


## K-means clustering
```{r}
km <- kmeans(boston_sc, centers = 4)
```

Well that was simple. Let's see what is the best number of clusters with a qplot. 
 library(ggplot2)
```{r}
# K-means might produce different results every time, because it randomly assigns the initial cluster centers. The function set.seed() can be used to deal with that.
set.seed(123)
# set the max number of clusters
k_max <- 10
# count the WCSS = within cluster sum of squares and plot it
twcss <- sapply(1:k_max, function(k){kmeans(boston_sc, k)$tot.withinss})
qplot(x = 1:k_max, y = twcss, geom = 'line')
```




To find the optimal number off clusters I used WCSS method. It tells how close the observations are to the cluster center. From the above graph the optimum number of clusters should be 2 (there is sudden drop from 2, see x-axis). 

## Visualize the clusters

```{r}
boston_sc <- as.data.frame(boston_sc)
# run kmeans again with 2 centroids
km <- kmeans(boston_sc, centers = 2)
# visualize the clusters
pairs(boston_sc[5:8], col = km$cluster)
```

since we used 2 to center our dataset, there are two clusters and the plotted with pairs. From the plot we can see for example, that the houses in the 1st cluster (in red) have
- lower AGE and lower NOX (nitrogen oxides concentration) 
- more rooms (RM, average number of rooms per dwelling), and higher distance to employment centers (DIS) 
