---
title: "chapter_5"
author: "Feroze"
date: "1 12 2019"
output: html_document
---

# Chapter 5: Dimensionality reduction techniques {.tabset}


we are continuing with chapter 4 the "human" dataset and do further analysis. From the previous exercises we know that this data includes human development indicators (columns) from countries (rows).
See more: http://hdr.undp.org/en/content/human-development-index-hdi

## loading and cleaning the data
```{r}
library(stringr)
library(dplyr)

human <- read.csv("http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/human1.txt", stringsAsFactors = F)

str(human$GNI)
human <- mutate(human, GNI = str_replace(human$GNI, pattern=",", replace ="") %>% as.numeric)
str(human$GNI)
```

#Excluding the unwanted variables
```{r}
names(human)
keep <- c("Country", "Edu2.FM", "Labo.FM", "Life.Exp", "Edu.Exp", "GNI", "Mat.Mor", "Ado.Birth", "Parli.F")
human <- dplyr::select(human, one_of(keep))
```

#Keeping only the vaules which are 'TRUE' and removing all rows with missing values i.e 'FALSE'
```{r}
complete.cases(human)
data.frame(human [-1], comp=complete.cases(human))
human_ <- filter(human, complete.cases(human))
complete.cases(human_)
```

#Removing the data relate to regions instead of countries
```{r}
dim(human_)
tail(human_, n=10)
last <- nrow(human_) - 7
human_ <- human_[1:last,]
dim(human_)
```

#Define the row names of the data and remove the country name column from the data.
```{r}
rownames(human_) <- human_$Country
human_ <- dplyr::select(human_, -Country)
dim(human_)
str(human_)

```

## Graphical overview and summaries of the variables
```{r}
library(GGally); library(ggplot2); library(corrplot)
ggpairs(human_)
cor(human_)%>%corrplot()
```
#We performed a correlation between two variables, the most strongest negative correlation is between mater mortality ratio and life expectancy at birth which is -0.4, and other negative correlations like life experience and Mater mortality. 
#there is no strong positive correlation, but slight correlation.

## Analysis of PCA without standardized data 
```{r}
pca_human <- prcomp(human_)
biplot(pca_human, choices = 1:2, cex = c(0.8,1), col = c("grey40", "deeppink2"))
```

## Standardize the variables in the human data and same analysis as above
```{r}
human_std <- scale(human_)
pca_human2 <- prcomp(human_std)
biplot(pca_human2, choices = 1:2, cex = c(0.8,1), col = c("grey40", "deeppink2"))
```
## Exploring the tea data and analyzing 
```{r}
library(ggplot2); library(FactoMineR); library(dplyr); library(tidyr)
data(tea)
str(tea)
dim(tea)
keep_columns <- c("Tea", "How", "how", "sugar", "where", "lunch")
tea_time <- dplyr::select(tea, one_of(keep_columns))
summary(tea_time)
str(tea_time)
```

## Visualizing the dataset and multiple correspondence analysis
```{r}
gather(tea_time) %>% ggplot(aes(value)) + facet_wrap("key", scales = "free") + geom_bar() + theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))
mac <- MCA(tea_time, graph = FALSE)
plot(mac, invisible=c("ind"), habillage = "quali")
```